Incremental tactics

### 2012-05-28 doc/papers/LFMTP2012/paper.tex
paper

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
\eg\ modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator.

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (type inference), but is also true in many
cases for what is usually called ``type checking'', or
``church-style'' calculi. Take for example dependent type systems,
relying on a certain notion of definitional equality on types, or type
systems with subtype polymorphism: both rely on a rule of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. It is of course the most
satisfactory because it allows to ``forget'' about the declarative
system and take only the result of the algorithm into consideration,
although these proofs can be difficult to obtain. But for the sake of
the discussion, let us choose another path: derivations in the
declarative systems, if they cannot always be algorithmically
inferred, can be \emph{verified} if provided by an oracle. This way,
even if the oracle is wrong, we have a way to decide it. Let us then
imagine a type-checker to be the pair of 1/ an untrusted procedure
$\finfer M$ for inferring a typing derivations out of programs $M$; it
doesn't have to be complete, or even sound, and 2/ a trusted kernel
$\fcheck{\md}$ for verifying these derivations in a given set of
inference rules $\Sigma$. This scheme has the following advantages
over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it is certifying: if $\fcheck{\finfer M} = \cst{true}$ then
  there really \emph{is} a derivation $\md$ for $M$. Of course the
  converse is not necessarily true: function $\finfer{}\!\!$ could
  fail, or have bugs (return an incorrect derivation).
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: %TODO third-party
                                              %verification
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\item \ldots\ but there is more:
\end{itemize}

@@

Compilers are generally monolithic programs, taking source code and
producing machine code in a series of transformations, one of the
first being type checking. But as type systems become richer, it
becomes increasingly difficult for the programmer to write a correct
program in one try, compile it and run it: writing a program longer
than a few dozen of lines becomes a tight and non-linear interaction
between the human and the type-checker involving constant experiments,
fixes, etc. If this interaction is possible for short programs and
fast checkers, it becomes increasingly difficult %TODO

>>>>

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
--- modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator.

@@

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (``type inference''), in which case the
missing information has to be algorithmically synthesized, but is also
true in many cases for what is commonly refered as ``type checking'',
or explicit ``church-style'' calculi. Take for example dependent type
systems, relying on a certain notion of definitional equality on
types, or type systems with subtype polymorphism: both rely on a rule
of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. Although these proofs
can be difficult to obtain, it is the most satisfactory because it
allows to ``forget'' about the declarative system and take only the
result of the algorithm into consideration. But for the sake of the
discussion, let us choose another path: derivations in the declarative
systems, if they cannot always be easily inferred, can be
\emph{verified} if provided by an oracle. Even if the oracle is wrong,
we can tell cheaply by verifying the witness typing derivation it
provides. Let us then imagine a type-checker to be the pair of 1/ an
untrusted procedure $\finfer M$ for inferring a typing derivations out
of programs $M$; it doesn't have to be complete, or even sound, and 2/
a trusted and fast kernel $\fcheck{\md}$ for verifying these
derivations in a given set of inference rules $\Sigma$. This is a
certifying scheme: if $\fcheck{\finfer M} = \cst{true}$ then there
really \emph{is} a derivation $\md$ for $M$. Of course the converse is
not necessarily true: function $\finfer{}\!\!$ could fail, or have
bugs (return an incorrect derivation). Explicitly generating the
derivation in the declarative system has the following advantages over
the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: certificates can be
  transmitted with the program, saving regeneration time and avoiding
  to disclose the generation program.
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\end{itemize}
\ldots\ but there is more:

@@

Compilers are generally monolithic programs, taking source code and
producing machine code in a series of transformations, one of the
first being type checking. But as type systems become richer, it
becomes increasingly difficult for the programmer to write a correct
program in one try, compile it and run it: writing a program longer
than a few dozen of lines becomes a tight and non-linear interaction
between the human and the type-checker involving constant experiments,
fixes, etc. If this interaction is possible for short programs and
fast checkers, it becomes increasingly hindered by the latency of
complete rechecks: even if the modifications made to the program
between two interactions are small, the whole program will be
rechecked. This is especially true for languages embedding formal
verification aspects, like proof assistants: there, tight interaction
with a type or proof checker is unavoidable due to the difficulty of
writing correct proofs without guidance from the system, and the time
taken by the tool to infer trivial parts of the proofs (proof
search). The constant modification of the source makes it necessary to
observe \emph{incrementally} the effect of a small change on the whole
edifice: we need to take advantage of the results of previous checks
in order to build the result for a modified version of the program.
